---
title: "NicheNet"
author: "Estefania"
date: "2024-07-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='Figs/', message=FALSE)
```

Installation

```{r}
devtools::install_github("saeyslab/nichenetr")
```

0. Load needed packages
```{r}

library("nichenetr")
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(dplyr)
library(here)
```

1. Creating liver Seurat object from the Tabula Muris file "Liver-counts.csv"

```{r}

    # Read the count matrix from a CSV file
    count_matrix <- read.csv("Liver-counts.csv", row.names = 1)

    # Create a Seurat object from the count matrix
    seurat_obj <- CreateSeuratObject(counts = count_matrix)
    seurat_obj

    # Normalize the data
    seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)

    # Add the annotations to the Seurat object ------------------------------------
    # Load annotations
    annot <- read.table("annotations_facs.csv", header = TRUE, sep = ",")

    # Create a temporary dataframe with the 'cell' column
    seurat_metadata <- seurat_obj@meta.data
    seurat_metadata$cell <- rownames(seurat_metadata)

    # Merge the metadata with the 'annot' dataframe based on the 'cell' column
    merged_metadata <- merge(seurat_metadata, annot[, c("cell", "cell_ontology_class")], by = "cell", all.x = TRUE)

    # Set the rownames of the merged metadata to be the cell names again
    rownames(merged_metadata) <- merged_metadata$cell
    merged_metadata$cell <- NULL  # Remove the temporary 'cell' column

    # Update the Seurat object's metadata
    seurat_obj@meta.data <- merged_metadata

    # Remove the dataframe
    rm(annot, merged_metadata, seurat_metadata)
    invisible(gc())
```
2. Giving Idents and other features to created Seurat Object

```{r}

all.genes <- rownames(seurat_obj)

Idents(object = seurat_obj) <- seurat_obj@meta.data$cell_ontology_class
seurat_obj <- ScaleData(seurat_obj, features = all.genes)

seurat_obj <- FindVariableFeatures(object = seurat_obj)
seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj))
seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)

```

3. Quality check (Seurat)

Exploring subset

```{r}
seurat_obj@meta.data$cell_ontology_class %>% table() 
```

```{r}
# Visualize QC metrics as a violin plot
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
DimPlot(seurat_obj, reduction = "umap")

```

4. Filtering out poor quality features


A. Features filtering. At least 10 counts and present in at least 5% of the cell type

```{r}
seurat_subset <- subset(seurat_obj, idents = c("Kupffer cell", "hepatocyte"))

# Access the raw data matrix from the Seurat object
data_matrix <- GetAssayData(seurat_subset, slot = "counts")

# Extract cell identities (cell types)
cell_types <- Idents(seurat_subset)

# Initialize a list to hold genes that pass the filters for both cell types
filtered_genes <- c()

# Loop through the cell types (e.g., "CellTypeA", "CellTypeB")
for (cell_type in unique(cell_types)) {
    
    # Subset the data matrix for cells of the current cell type
    cell_type_cells <- WhichCells(seurat_subset, idents = cell_type)
    cell_type_data <- data_matrix[, cell_type_cells]
    
    # Step 1: Filter for genes with at least 10 counts in any cell
    genes_with_min_counts <- rowSums(cell_type_data >= 10) > 0
    
    # Step 2: Filter for genes present in at least 5% of cells within this cell type
    min_cells <- length(cell_type_cells) * 0.05  # Calculate 5% of cells for this type
    genes_in_min_cells <- rowSums(cell_type_data > 0) >= min_cells
    
    # Combine both conditions
    valid_genes <- rownames(cell_type_data)[genes_with_min_counts & genes_in_min_cells]
    
    # Store valid genes for this cell type
    filtered_genes <- union(filtered_genes, valid_genes)
}

```

B. Creating new Seurat Object with the subset
```{r}
counts <- GetAssayData(seurat_obj, slot="counts", assay="RNA")   
#genes.percent.expression <- rowMeans(counts>0 )*100   

#genes.filter <- names(genes.percent.expression[genes.percent.expression>5])  #select genes expressed in at least 1% of cells
counts.sub <- counts[filtered_genes,]
new_seurat_object <- CreateSeuratObject(counts=counts.sub)
```

C. Normalizing and adding metadata to the new seurat object with the subset of genes (filtered_genes)
```{r}
# Normalize the data
    new_seurat_object <- NormalizeData(new_seurat_object, normalization.method = "LogNormalize", scale.factor = 10000)

    # Add the annotations to the Seurat object ------------------------------------
    # Load annotations
    annot <- read.table("annotations_facs.csv", header = TRUE, sep = ",")

    # Create a temporary dataframe with the 'cell' column
    seurat_metadata <- new_seurat_object@meta.data
    seurat_metadata$cell <- rownames(seurat_metadata)

    # Merge the metadata with the 'annot' dataframe based on the 'cell' column
    merged_metadata <- merge(seurat_metadata, annot[, c("cell", "cell_ontology_class")], by = "cell", all.x = TRUE)

    # Set the rownames of the merged metadata to be the cell names again
    rownames(merged_metadata) <- merged_metadata$cell
    merged_metadata$cell <- NULL  # Remove the temporary 'cell' column

    # Update the Seurat object's metadata
    new_seurat_object@meta.data <- merged_metadata

    # Remove the dataframe
    rm(annot, merged_metadata, seurat_metadata)
    invisible(gc())
```
D. Add Idents to new Seurat object
```{r}

all.genes <- rownames(new_seurat_object)
Idents(object = new_seurat_object) <- new_seurat_object@meta.data$cell_ontology_class

new_seurat_object <- ScaleData(new_seurat_object, features = all.genes)

new_seurat_object <- FindVariableFeatures(object = new_seurat_object)
new_seurat_object <- RunPCA(new_seurat_object, features = VariableFeatures(object = new_seurat_object))
new_seurat_object <- RunUMAP(new_seurat_object, dims = 1:10)

```

E. Quality control new Seurat ibject with filtered features

Exploring subset

```{r}
new_seurat_object@meta.data$cell_ontology_class %>% table() 
```

```{r}
# Visualize QC metrics as a violin plot
VlnPlot(new_seurat_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
DimPlot(new_seurat_object, reduction = "umap")

```

```{r}
FeaturePlot(new_seurat_object, features = "Hnf4a", reduction = "umap")
```


5. Subseting cell populations of interest (Kupffer cells and hepatocytes)

```{r}
seurat_obj <- subset(new_seurat_object,
        subset = cell_ontology_class %in% c("Kupffer cell", "hepatocyte"))

Idents(object = seurat_obj) <- seurat_obj@meta.data$cell_ontology_class
```
Exploring subset

```{r}
seurat_obj@meta.data$cell_ontology_class %>% table() 
```

6. Loading Nichenet and EVice data from zenodo
Note for FNia: change ligand_to_target and add the zenodo address 
```{r}
organism <- "mouse"

if(organism == "human"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
  #ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
} else if(organism == "mouse"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
  #ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))

}

lr_network <- lr_network %>% distinct(from, to)
head(lr_network)
```


7. Define receiver cell populations (KC and hepatocytes) and get features present in both
```{r}
# Step 1: Extract the cells belonging to each cell type
cells_type_1 <- WhichCells(new_seurat_object, idents = "Kupffer cell")
cells_type_2 <- WhichCells(new_seurat_object, idents = "hepatocyte")

# Step 2: Subset the Seurat object to get the count matrix for each cell type
data_matrix <- GetAssayData(new_seurat_object, slot = "counts")

# Get the features (genes) expressed in each cell type
# Features expressed in at least one cell for cell_type_1
features_type_1 <- rownames(data_matrix)[rowSums(data_matrix[, cells_type_1] > 0) > 0]

# Features expressed in at least one cell for cell_type_2
features_type_2 <- rownames(data_matrix)[rowSums(data_matrix[, cells_type_2] > 0) > 0]

# Step 3: Combine the two lists of features and remove duplicates
combined_features <- unique(c(features_type_1, features_type_2))

# Output the combined list of features
#combined_features
```


```{r}
receiver = c("Kupffer cell", "hepatocyte")
#expressed_genes_receiver <- get_expressed_genes(receiver,  new_seurat_object@assays[["RNA"]], new_seurat_object@meta.data[["cell_ontology_class"]], pct = 0.00)

#expressed_genes_receiver <- rownames(new_seurat_object@assays[["RNA"]])

expressed_genes_receiver <- combined_features
```

8. Get list of all proteins (call ligands, so this should be changed)

```{r}
gr_network <- readRDS(url("https://zenodo.org/records/7074291/files/gr_network_mouse_21122021.rds"))

ligands <- unique(c(weighted_networks[["lr_sig"]][["from"]], weighted_networks[["lr_sig"]][["to"]], gr_network$from))
```

9. Transpose ligand-to-target matrix (maybe do it before?)

```{r}
ligand_to_target <- t(ligand_to_target)
```


10. Defining potential proteins as all proteins whose interacting proteins (receptors, intermediary signalling proteins and tf) are expressed in the receiving cell populations
```{r}
all_receptors <- unique(rownames(ligand_to_target))  
expressed_receptors <- intersect(all_receptors, expressed_genes_receiver)

lr_sig <- weighted_networks[["lr_sig"]]
gr <- weighted_networks[["gr"]]
potential_ligands <- lr_sig[lr_sig$to %in% expressed_receptors, "from"]
potential_ligands2 <- gr[gr$to %in% expressed_receptors, "from"]
potential_ligands <- unique(c(potential_ligands$from, potential_ligands2$from))
```

New chunk new nomencalture DO NOT run until changing all Nomenclature

```{r}
all_genes <- unique(rownames(EV_cargo_target_matrix))  
expressed_interactors <- intersect(all_genes, expressed_genes_receiver)

lr_sig <- weighted_networks[["lr_sig"]]
gr <- weighted_networks[["gr"]]

potential_EV_cargo_prot <- lr_sig[lr_sig$to %in% expressed_interactors, "from"]
potential_EV_cargo_tf <- gr[gr$to %in% expressed_interactors, "from"]
potential_EV_cargo <- unique(c(potential_EV_cargo_prot$from, potential_EV_cargo_tf$from))
```


11. Upload differentially abundant proteins of interest present i the EVs data (HFD condition in this example)

```{r}

```

12. Get list of potential proteins for sender

```{r}
potential_ligands <- intersect(potential_ligands, HFD_ligands)
```

MAYBE DELETE This
Potential ligands as all ligands whose cognate receptors are expressed - Modification: using only proteins present in EVs (HFD)
Run for HFD
```{r}
all_receptors <- unique(rownames(ligand_to_target))  
expressed_receptors <- intersect(all_receptors, expressed_genes_receiver)

lr_sig <- weighted_networks[["lr_sig"]]
gr <- weighted_networks[["gr"]]
potential_ligands <- lr_sig[lr_sig$to %in% expressed_receptors, "from"]
potential_ligands2 <- gr[gr$to %in% expressed_receptors, "from"]
potential_ligands <- unique(c(potential_ligands$from, potential_ligands2$from))

potential_ligands <- intersect(potential_ligands, HFD_ligands)
```

Potential ligands as all ligands whose cognate receptors are expressed - Modification: using only proteins present in EVs (NCD)
```{r}
all_receptors <- unique(rownames(ligand_to_target))  
expressed_receptors <- intersect(all_receptors, expressed_genes_receiver)

lr_sig <- weighted_networks[["lr_sig"]]
gr <- weighted_networks[["gr"]]
potential_ligands <- lr_sig[lr_sig$to %in% expressed_receptors, "from"]
potential_ligands2 <- gr[gr$to %in% expressed_receptors, "from"]
potential_ligands <- unique(c(potential_ligands$from, potential_ligands2$from))

EV_cargo <- intersect(EV_cargo, 
                      potential_EV_cargo)
```
UNTIL HERE

13. Define the gene set of interest
Note: We had to add an identity class label with Idents

```{r}
condition_oi <-  "Kupffer cell"
condition_reference <- "hepatocyte"

#Idents(object = seurat_obj) <- seurat_obj@meta.data$celltype

seurat_obj_receiver <- subset(seurat_obj, idents = receiver)

DE_table_receiver <-  FindMarkers(object = seurat_obj,
                                  ident.1 = condition_oi, ident.2 = condition_reference,
                                  group.by = "cell_ontology_class",
                                  min.pct = 0.05) %>% rownames_to_column("gene")

geneset_oi <- DE_table_receiver %>% filter(p_val_adj <= 0.05 & avg_log2FC >= 0.25) %>% pull(gene)
geneset_oi <- geneset_oi %>% .[. %in% rownames(ligand_to_target)]
```

14. Defining background genes

```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_to_target)]
```

15. Perform protein activity analysis

```{r}
ligand_activities <- predict_ligand_activities(geneset = geneset_oi,
                                               background_expressed_genes = background_expressed_genes,
                                               ligand_target_matrix = big_ligand_to_target,
                                               potential_ligands = potential_ligands)

ligand_activities <- ligand_activities %>% arrange(-aupr_corrected) %>% mutate(rank = rank(desc(aupr_corrected)))
ligand_activities

```
16. Visualization of top-ranked ligands

```{r}
p_hist_lig_activity <- ggplot(ligand_activities, aes(x=aupr_corrected)) + 
  geom_histogram(color="black", fill="darkorange")  + 
  geom_vline(aes(xintercept=min(ligand_activities %>% top_n(30, aupr_corrected) %>% pull(aupr_corrected))),
             color="red", linetype="dashed", size=1) + 
  labs(x="ligand activity (PCC)", y = "# ligands") +
  theme_classic()

p_hist_lig_activity
```
We can also visualize the ligand activity measure (AUPR) of these top-ranked ligands:
changed to top 100

```{r}
best_upstream_ligands <- ligand_activities %>% top_n(30, aupr_corrected) %>% arrange(-aupr_corrected) %>% pull(test_ligand)

vis_ligand_aupr <- ligand_activities %>% filter(test_ligand %in% best_upstream_ligands) %>%
  column_to_rownames("test_ligand") %>% select(aupr_corrected) %>% arrange(aupr_corrected) %>% as.matrix(ncol = 1)

(make_heatmap_ggplot(vis_ligand_aupr,
                     "Prioritized ligands", "Ligand activity", 
                     legend_title = "AUPR", color = "darkorange") + 
    theme(axis.text.x.top = element_blank()))  
```

17. Infer target genes  of top-ranked proteins

```{r}
active_ligand_target_links_df <- best_upstream_ligands %>%
  lapply(get_weighted_ligand_target_links,
         geneset = geneset_oi,
         ligand_target_matrix = ligand_to_target,
         n = 100) %>%
  bind_rows() %>% drop_na()
```

```{r}
active_ligand_target_links <- prepare_ligand_target_visualization(
  ligand_target_df = active_ligand_target_links_df,
  ligand_target_matrix = ligand_to_target,
  cutoff = 0.3) 
```

```{r}
order_ligands <- intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
order_targets <- active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links))

vis_ligand_target <- t(active_ligand_target_links[order_targets,order_ligands])

target_genes_heatmap <- make_heatmap_ggplot(vis_ligand_target, "Prioritized ligands", "Predicted target genes",
                    color = "purple", legend_title = "Regulatory potential") +
  scale_fill_gradient2(low = "whitesmoke",  high = "purple")
```

```{r}
target_genes_heatmap
```

```{r}
png("target_genes_heatmap_HFD_hyperparameters_avg_26_02_25.png", res = 300, width = 5000, height = 2000)
print(target_genes_heatmap)
```

Old code, cant be run, needs to be modified. Seem ok now 18/10

```{r}

lr_network <- active_ligand_target_links_df
lr_network$weight <- NULL
lr_network$from <- lr_network$ligand
lr_network$to <- lr_network$target

ligand_receptor_links_df <- get_weighted_ligand_receptor_links(
  best_upstream_ligands, expressed_receptors,
  lr_network, weighted_networks$lr_sig)

ligand_receptor_links_df2 <- get_weighted_ligand_receptor_links(
  best_upstream_ligands, expressed_receptors,
  lr_network, weighted_networks$gr)

ligand_receptor_links_df3 <- ligand_receptor_links_df2[1:2,]

#test proteins for EVs
ligand_receptor_links_df4 <- get_weighted_ligand_receptor_links(
  ligand_activities$test_ligand, expressed_receptors,
  lr_network, weighted_networks$lr_sig)

test <- unique(ligand_receptor_links_df4$to)
```

Receptors of top-ranked ligands

```{r}

vis_ligand_receptor_network <- prepare_ligand_receptor_visualization(
  ligand_receptor_links_df,
  best_upstream_ligands,
  order_hclust = "both") 

heatmap_top30 <- make_heatmap_ggplot(t(vis_ligand_receptor_network), 
                     y_name = "Ligands", x_name = "Receptors",  
                     color = "mediumvioletred", legend_title = "Prior interaction potential")
```

```{r}
heatmap_top30
```

```{r}
png("top30heatmap.png", res = 300, width = 2000, height = 4400)
print(heatmap_top30)
```

Ligands from the EVs (load data from ComparativeTreeplots_EMBO)

```{r}

potential_ligands_NCD <- intersect(ligands, NCD_ligands)
potential_ligands_HFD <- intersect(ligands, HFD_ligands)
```

Loading zenodo data for signalling pathways
Note for FNia: Check if these datasets were already loaded
```{r}
weighted_networks <- readRDS(url("https://zenodo.org/records/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))
ligand_tf_matrix <- readRDS(url("https://zenodo.org/records/7074291/files/ligand_tf_matrix_nsga2r_final_mouse.rds"))

lr_network <- readRDS(url("https://zenodo.org/records/7074291/files/lr_network_mouse_21122021.rds"))
sig_network <- readRDS(url("https://zenodo.org/records/7074291/files/signaling_network_mouse_21122021.rds"))
gr_network <- readRDS(url("https://zenodo.org/records/7074291/files/gr_network_mouse_21122021.rds"))
```

Intersect 100 prioritized ligands and potential ligands NCD/HFD

```{r}
proteins_HFD <- intersect(ligand_activities$test_ligand, potential_ligands_HFD)
```


Inferring ligand-to-target signaling paths NCD

```{r}
ligands_oi <- "Apoe"
targets_oi <- c("Il1b", "Cd320", "Pon2")

active_signaling_network <- get_ligand_signaling_path(ligands_all = ligands_oi,
                                                      targets_all = targets_oi, 
                                                      weighted_networks = weighted_networks,
                                                      ligand_tf_matrix = ligand_to_target,
                                                      top_n_regulators = 4,
                                                      minmax_scaling = TRUE) 


graph_min_max <- diagrammer_format_signaling_graph(signaling_graph_list = active_signaling_network,
                                                   ligands_all = ligands_oi, targets_all = targets_oi,
                                                   sig_color = "indianred", gr_color = "steelblue")

# To render the graph in RStudio Viewer, uncomment following line of code
 DiagrammeR::render_graph(graph_min_max, layout = "tree")

# To export/draw the svg, you need to install DiagrammeRsvg
#graph_svg <- DiagrammeRsvg::export_svg(DiagrammeR::render_graph(graph_min_max, layout = "tree", output = "graph"))
#cowplot::ggdraw() + cowplot::draw_image(charToRaw(graph_svg))
```

Intersect 100 prioritized ligands and potential ligands NCD

```{r}
proteins_NCD <- intersect(ligand_activities$test_ligand, potential_ligands_NCD)
```


Inferring ligand-to-target signaling paths HFD

```{r}
ligands_oi <- "Scly" # this can be a list of multiple ligands if required
targets_oi <- c("Apbb1ip", "Ier5", "Il1r1", "Laptm5", "Ly86", "Pik3r5", "Prkcb", "Cxcr4", "Lilra5", "Lilra6", "Pira2", "Tbc1d1", "Tjp2", "Cd40", "Cd83", "Mmp9", "Pmaip1", "Pira6", "Siglec1", "Timd4", "Cotl1", "Frmd4a", "Odc1", "Spn", "Grb2", "Csf1r", "Ctsb", "Cd14", "Plac8", "H3f3b", "Lmo2", "Alox5ap", "Anxa3", "Foxo1", "Zeb2")

active_signaling_network <- get_ligand_signaling_path(ligands_all = ligands_oi,
                                                      targets_all = targets_oi, 
                                                      weighted_networks = weighted_networks,
                                                      ligand_tf_matrix = ligand_to_target,
                                                      top_n_regulators = 4,
                                                      minmax_scaling = TRUE) 


#graph_min_max <- diagrammer_format_signaling_graph(signaling_graph_list = active_signaling_network,
                                                   #ligands_all = ligands_oi, targets_all = targets_oi,
                                                   #sig_color = "indianred", gr_color = "steelblue")

# To render the graph in RStudio Viewer, uncomment following line of code
# DiagrammeR::render_graph(graph_min_max, layout = "tree")

# To export/draw the svg, you need to install DiagrammeRsvg
#graph_svg <- DiagrammeRsvg::export_svg(DiagrammeR::render_graph(graph_min_max, layout = "tree", output = "graph"))
#cowplot::ggdraw() + cowplot::draw_image(charToRaw(graph_svg))
```


Substracting NCD EVs targets (signalling, regulatory)

```{r}
vector_targets_NCD <- active_signaling_network[["sig"]][["to"]]
vector_targets_NCD <- unique(vector_targets_NCD)
```

```{r}
vector_targets_reg_NCD <- active_signaling_network[["gr"]][["to"]]
vector_targets_reg_NCD <- unique(vector_targets_reg_NCD)
```


Cluster profiler analysis

```{r}
library("clusterProfiler")

library("ggplot2")

library("forcats")

library("ggstance")

library("org.Mm.eg.db")

library("DOSE")

library("ggridges")

library("ggupset")

library("enrichplot")

library("ggrepel")

library("ggnewscale")

library(cowplot)

library("pathview")

library("ReactomePA")

library(stringr)

library(dplyr)

library(limma)

library(R.utils)

library("GOSemSim")

library("ggtreeExtra")

options(clusterProfiler.download.method = "wininet")
```


Defining background genes

```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
```

Dotplot analysis
```{r}
ego <- enrichGO(gene = vector_targets_NCD,
 universe = background_expressed_genes,
 keyType = "SYMBOL",
 OrgDb = org.Mm.eg.db,
 ont = "BP",
 pAdjustMethod = "BH",
 qvalueCutoff = 0.05,
 readable = TRUE,
 minGSSize = 10,
  maxGSSize = 1000)
# Output results from GO analysis to a table
cluster_summary <- data.frame(ego)

# Make a dotplot
dotplot(ego, showCategory=20)
```

Inferring ligand-to-target signaling paths HFD (same names as in NCD, bear that in mind)

```{r}
ligands_oi <- potential_ligands # this can be a list of multiple ligands if required
targets_oi <- expressed_receptors

active_signaling_network <- get_ligand_signaling_path(ligands_all = ligands_oi,
                                                      targets_all = targets_oi, 
                                                      weighted_networks = weighted_networks,
                                                      ligand_tf_matrix = ligand_tf_matrix,
                                                      top_n_regulators = 4,
                                                      minmax_scaling = TRUE) 


#graph_min_max <- diagrammer_format_signaling_graph(signaling_graph_list = active_signaling_network,
                                                   #ligands_all = ligands_oi, targets_all = targets_oi,
                                                   #sig_color = "indianred", gr_color = "steelblue")

# To render the graph in RStudio Viewer, uncomment following line of code
# DiagrammeR::render_graph(graph_min_max, layout = "tree")

# To export/draw the svg, you need to install DiagrammeRsvg
#graph_svg <- DiagrammeRsvg::export_svg(DiagrammeR::render_graph(graph_min_max, layout = "tree", output = "graph"))
#cowplot::ggdraw() + cowplot::draw_image(charToRaw(graph_svg))
```

Substracting HFD EVs targets (signalling, regulatory)

```{r}
vector_targets_HFD <- active_signaling_network[["sig"]][["to"]]
vector_targets_HFD <- unique(vector_targets_HFD)
```

```{r}
vector_targets_reg_HFD <- active_signaling_network[["gr"]][["to"]]
vector_targets_reg_HFD <- unique(vector_targets_reg_HFD)
```


Visualizing network

```{r}
library(DiagrammeR)

DiagrammeR::render_graph(graph_min_max, layout = "tree")
```
Comparison clusterprofiler

Get unique genes for both groups.
Reg targets are all the same.
```{r}
unique_targets_NCD <- setdiff(vector_targets_NCD, vector_targets_HFD)
unique_targets_HFD <- setdiff(vector_targets_HFD, vector_targets_NCD)

unique_targets_reg_NCD <- setdiff(vector_targets_reg_NCD, vector_targets_reg_HFD)
unique_targets_reg_HFD <- setdiff(vector_targets_reg_HFD, vector_targets_reg_NCD)
```

```{r}
datafNCD <- data.frame(Genes = unique_targets_NCD)
datafNCD$group <- "NCD"

datafHFD <- data.frame(Genes = unique_targets_HFD)
datafHFD$group <- "HFD"
```

```{r}
entre_NCD = bitr(datafNCD$Genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")
datafNCD <- merge(datafNCD, entre_NCD , by.x = "Genes", by.y ="SYMBOL")
datafNCD$Genes <- NULL

entre_HFD = bitr(datafHFD$Genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")
datafHFD <- merge(datafHFD, entre_HFD , by.x = "Genes", by.y ="SYMBOL")
datafHFD$Genes <- NULL

comparison_data <- rbind(datafNCD, datafHFD)
comparison_data$group <- as.factor(comparison_data$group)
```
Comparison using Clusterprofiler

```{r}
#comparePlot <- compareCluster(OrgDb="org.Mm.eg.db", ENTREZID~group, data=comparison_data, fun="enrichGO", readable = TRUE)

 pp <- dotplot(comparePlot, x= ~group, by = "count", title = "HFD vs.NCD", showCategory = REVIGO_names) + aes(x=fct_relevel(group)) + xlab(NULL) +
 scale_color_gradientn(colours=c("#cc3300", "#ff9966", "#ffffcc"),
 guide=guide_colorbar(reverse=TRUE, order=1)) +
 guides(size = guide_legend(override.aes=list(shape=1))) +
 theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
 panel.grid.major.x = element_blank())


print(pp)

```

```{r}
png("MF_comparison.png", res = 300, width = 2000, height = 3500)
print(pp)
```

Uploding REVIGO list

```{r}
REVIGO_1 <- read.csv("C:/Users/tefi1/Documents/PhD/Data_analysis/EMBO/Revigo_1.csv", check.names = FALSE)

REVIGO_names <- REVIGO_1$REVIGO_1
```


```{r}

library(writexl)

comparePlotGO_tab=as.data.frame(comparePlot)

write_xlsx(
     comparePlotGO_tab,
     path = "C:/Users/tefi1/Documents/PhD/Data_analysis//EMBO/comparePlot.xlsx",
     col_names = TRUE,
     format_headers = TRUE,
     use_zip64 = FALSE)
```



Chunk make be used with a different dataset
```{r}
datafNCD <- data.frame(Genes = vector_targets_NCD)
datafNCD$type <- "signalling"

datafNCD2 <- data.frame(Genes = vector_targets_reg_NCD)
datafNCD2$type <- "regulatory"

datafNCD <- merge(datafNCD2, datafNCD , by = "Genes", all = TRUE)

datafNCD$type <- ifelse(!is.na(datafNCD$type.x) & !is.na(datafNCD$type.y), "both", 
                        ifelse(!is.na(datafNCD$type.x), "regulatory", "signalling"))
```


```{r}
datafHFD <- data.frame(Genes = vector_targets_HFD)
datafHFD$type <- "signalling"

datafHFD2 <- data.frame(Genes = vector_targets_reg_HFD)
datafHFD2$type <- "regulatory"

datafHFD <- merge(datafHFD2, datafHFD , by = "Genes", all = TRUE)

datafHFD$type <- ifelse(!is.na(datafHFD$type.x) & !is.na(datafHFD$type.y), "both", 
                        ifelse(!is.na(datafHFD$type.x), "regulatory", "signalling"))
```



Which of the collected data sources support the interactions in this network

```{r}
data_source_network <- infer_supporting_datasources(signaling_graph_list = active_signaling_network,
                                                    lr_network = lr_network, sig_network = sig_network, gr_network = gr_network)
head(data_source_network) 
```
Exporting to cytoscape

HFD
```{r}
output_path <- ""
write_output <- TRUE # change to TRUE for writing output

# weighted networks ('import network' in Cytoscape)
if(write_output){
  bind_rows(active_signaling_network$sig %>% mutate(layer = "signaling"),
            active_signaling_network$gr %>% mutate(layer = "regulatory")) %>%
    write_tsv(paste0(output_path,"weighted_signaling_network_HFD_potential_ligands_Scly.txt")) 
}

# networks with information of supporting data sources ('import network' in Cytoscape)
if(write_output){
  data_source_network %>% write_tsv(paste0(output_path,"data_source_network_HFD_potential_ligands_Scly.txt"))
}

# Node annotation table ('import table' in Cytoscape)
specific_annotation_tbl <- bind_rows(
  tibble(gene = ligands_oi, annotation = "ligand"),
  tibble(gene = targets_oi, annotation = "target"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,ligands_oi)) %>% intersect(lr_network$to %>% unique()), annotation = "receptor"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,ligands_oi)) %>% intersect(gr_network$from %>% unique()) %>% setdiff(c(data_source_network$from, data_source_network$to) %>% unique() %>% intersect(lr_network$to %>% unique())),annotation = "transcriptional regulator")
)
non_specific_annotation_tbl <- tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(specific_annotation_tbl$gene), annotation = "signaling mediator")

if(write_output){
  bind_rows(specific_annotation_tbl, non_specific_annotation_tbl) %>%
    write_tsv(paste0(output_path,"annotation_table_HFD_potential_ligands_Scly.txt"))
}
```

NCD
```{r}
output_path <- ""
write_output <- TRUE # change to TRUE for writing output

# weighted networks ('import network' in Cytoscape)
if(write_output){
  bind_rows(active_signaling_network$sig %>% mutate(layer = "signaling"),
            active_signaling_network$gr %>% mutate(layer = "regulatory")) %>%
    write_tsv(paste0(output_path,"weighted_signaling_network_NCD_potential_ligands_APOE.txt")) 
}

# networks with information of supporting data sources ('import network' in Cytoscape)
if(write_output){
  data_source_network %>% write_tsv(paste0(output_path,"data_source_network_NCD_potential_ligands_APOE.txt"))
}

# Node annotation table ('import table' in Cytoscape)
specific_annotation_tbl <- bind_rows(
  tibble(gene = ligands_oi, annotation = "ligand"),
  tibble(gene = targets_oi, annotation = "target"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,ligands_oi)) %>% intersect(lr_network$to %>% unique()), annotation = "receptor"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,ligands_oi)) %>% intersect(gr_network$from %>% unique()) %>% setdiff(c(data_source_network$from, data_source_network$to) %>% unique() %>% intersect(lr_network$to %>% unique())),annotation = "transcriptional regulator")
)
non_specific_annotation_tbl <- tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(specific_annotation_tbl$gene), annotation = "signaling mediator")

if(write_output){
  bind_rows(specific_annotation_tbl, non_specific_annotation_tbl) %>%
    write_tsv(paste0(output_path,"annotation_table_NCD_potential_ligands_APOE.txt"))
}
```

