---
title: "NicheNet's ligand activity analysis on a gene set of interest: predict active ligands and their target genes"
author: "Robin Browaeys"
date: "2019-01-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NicheNet's ligand activity analysis on a gene set of interest: predict active ligands and their target genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: library.bib
---

<!-- github markdown built using 
rmarkdown::render("vignettes/ligand_activity_geneset.Rmd", output_format = "github_document")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This vignette follows the steps described in [Perform NicheNet analysis starting from a Seurat object: step-by-step analysis](seurat_steps.md) with two major differences: a predefined gene set of interest is given, and a different definition of expressed genes.

Here, we use explore intercellular communication in the tumor microenvironment of head and neck squamous cell carcinoma (HNSCC) [@puram_single-cell_2017]. More specifically, we will look at which ligands expressed by cancer-associated fibroblasts (CAFs) can induce a specific gene program in neighboring malignant cells. The original authors of the study have linked this partial epithelial-mesenschymal transition (p-EMT) program to metastasis.

The used [ligand-target matrix](https://doi.org/10.5281/zenodo.7074290) and example [expression data](https://doi.org/10.5281/zenodo.3260758) of interacting cells can be downloaded from Zenodo.

# Prepare NicheNet analysis

### Load packages

```{r}
library(nichenetr)
library(tidyverse)
```

### Read in NicheNet's networks

The ligand-target prior model, ligand-receptor network, and weighted integrated networks are needed for this vignette. The ligand-target prior model is a matrix describing the potential that a ligand may regulate a target gene, and it is used to run the ligand activity analysis. The ligand-receptor network contains information on potential ligand-receptor bindings, and it is used to identify potential ligands. Finally, the weighted ligand-receptor network contains weights representing the potential that a ligand will bind to a receptor, and it is used for visualization.

```{r}
organism <- "human"

if(organism == "human"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
  ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
} else if(organism == "mouse"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
  ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))

}

lr_network <- lr_network %>% distinct(from, to)
head(lr_network)
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns

head(weighted_networks$lr_sig) # interactions and their weights in the ligand-receptor + signaling network
head(weighted_networks$gr) # interactions and their weights in the gene regulatory network

```

### Read in the expression data of interacting cells

This is publicly available single-cell data from CAF and malignant cells from HNSCC tumors.

```{r}
hnscc_expression <- readRDS(url("https://zenodo.org/record/3260758/files/hnscc_expression.rds"))
expression <- hnscc_expression$expression
sample_info <- hnscc_expression$sample_info # contains meta-information about the cells
```

Because the NicheNet v2 networks are in the most recent version of the official gene symbols, we will make sure that the gene symbols used in the expression data are also updated (= converted from their “aliases” to official gene symbols).

```{r}
# If this is not done, there will be 35 genes fewer in lr_network_expressed!
colnames(expression) <- convert_alias_to_symbols(colnames(expression), "human", verbose = FALSE)
```

## 1. Define a set of potential ligands

Our research question is to prioritize which ligands expressed by CAFs can induce p-EMT in neighboring malignant cells. Hence, we will only use on the **sender-focused** approach, with CAFs as senders and malignant cells as receivers.

The set of potential ligands is defined as ligands that are expressed in sender cells whose cognate receptors are also expressed in receiver cells.

So first, we will determine which genes are expressed in the sender cells (CAFs) and receiver cells (malignant cells). We will only consider samples from high quality primary tumors and also remove samples from lymph node metastases. We will use the definition of expressed genes by the original authors, that is, the aggregate expression of each gene $i$ across the $k$ cells, calculated as $E_a(i) = log_{2}(average(TPM(i)1…k)+1)$, should be >= 4.

We recommend users to define expressed genes in the way that they consider to be most appropriate for their dataset. For single-cell data generated by the 10x platform in our lab, we consider genes to be expressed in a cell type when they have non-zero values in a certain fraction of the cells from that cell type (usually 10%). This is used in the vignette [Perform NicheNet analysis starting from a Seurat object: step-by-step analysis](seurat_steps.md).

```{r}
tumors_remove <- c("HN10","HN","HN12", "HN13", "HN24", "HN7", "HN8","HN23")

CAF_ids <- sample_info %>%
  filter(`Lymph node` == 0 & !(tumor %in% tumors_remove) &
           `non-cancer cell type` == "CAF") %>% pull(cell)
malignant_ids <- sample_info %>% filter(`Lymph node` == 0 &
                                          !(tumor %in% tumors_remove) &
                                          `classified  as cancer cell` == 1) %>% pull(cell)

expressed_genes_sender <- expression[CAF_ids,] %>%
  apply(2,function(x){10*(2**x - 1)}) %>%
  apply(2,function(x){log2(mean(x) + 1)}) %>% .[. >= 4] %>% 
  names()

expressed_genes_receiver <- expression[malignant_ids,] %>%
  apply(2,function(x){10*(2**x - 1)}) %>%
  apply(2,function(x){log2(mean(x) + 1)}) %>% .[. >= 4] %>%
  names()

length(expressed_genes_sender)
length(expressed_genes_receiver)
```

Now, we can filter the expressed ligands and receptors to only those that putatively bind together. This information is stored in NicheNet's ligand-receptor network by gathering various data sources.

```{r}
ligands <- lr_network %>% pull(from) %>% unique()
expressed_ligands <- intersect(ligands,expressed_genes_sender)

receptors <- lr_network %>% pull(to) %>% unique()
expressed_receptors <- intersect(receptors,expressed_genes_receiver)

potential_ligands <-  lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% 
  pull(from) %>% unique()

head(potential_ligands)
```

## 2. Define the gene set of interest and a background of genes

The gene set of interest consists of genes for which the expression is possibly affected due to communication with other cells. The definition of this gene set depends on your research question and is a crucial step in the use of NicheNet.

Here, we will use the p-EMT gene set defined by the original authors as gene set of interest to investigate how CAFs can induce p-EMT in malignant cells.

```{r}
# Only consider genes also present in the NicheNet model - this excludes genes from the gene list for which the official HGNC symbol was not used by Puram et al.
geneset_oi <- readr::read_tsv(url("https://zenodo.org/record/3260758/files/pemt_signature.txt"),
                              col_names = "gene") %>%
  pull(gene) %>% .[. %in% rownames(ligand_target_matrix)] 

length(geneset_oi)
```

## 3. Define background genes

We will all genes expressed in malignant cells as the background set.

```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]

length(background_expressed_genes)
```

## 4. Perform NicheNet ligand activity analysis

With the ligand activity analysis, we assess how well each CAF-ligand can predict the p-EMT gene set compared to the background of expressed genes.

```{r}
ligand_activities <- predict_ligand_activities(geneset = geneset_oi,
                                               background_expressed_genes = background_expressed_genes,
                                               ligand_target_matrix = ligand_target_matrix,
                                               potential_ligands = potential_ligands)

```

Ligands are ranked based on the area under the precision-recall curve (AUPR) between a ligand's target predictions and the observed transcriptional response. Although other metrics like the AUROC and pearson correlation coefficient are also computed, we demonstrated in our validation study that the AUPR was the most informative measure to define ligand activity (this was the Pearson correlation for v1). The vignette on how we performed the validation can be found at [Evaluation of NicheNet's ligand-target predictions](model_evaluation.md).

```{r}
(ligand_activities <- ligand_activities %>% arrange(-aupr_corrected) %>%
  mutate(rank = rank(desc(aupr_corrected))))

best_upstream_ligands <- ligand_activities %>% top_n(30, aupr_corrected) %>%
  arrange(-aupr_corrected) %>% pull(test_ligand)

best_upstream_ligands
```
We will use the top 30 ligands to predict active target genes and construct an active ligand-receptor network. 

## 5. Infer target genes and receptors of top-ranked ligands

Active target genes are defined as genes in the gene set of interest that have the highest regulatory potential for each top-ranked ligand. These top targets of each ligand are based on the prior model. Specifically, the function get_weighted_ligand_target_links will return genes that are in the gene set of interest and are the top `n` targets of a ligand (default: `n = 200`).

```{r ligand-target-heatmap, fig.width=8, fig.height=6}
active_ligand_target_links_df <- best_upstream_ligands %>%
  lapply(get_weighted_ligand_target_links,
         geneset = geneset_oi,
         ligand_target_matrix = ligand_target_matrix,
         n = 200) %>% bind_rows()

active_ligand_target_links <- prepare_ligand_target_visualization(
  ligand_target_df = active_ligand_target_links_df,
  ligand_target_matrix = ligand_target_matrix,
  cutoff = 0.25)

order_ligands <- intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
order_targets <- active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links))

vis_ligand_target <- t(active_ligand_target_links[order_targets,order_ligands])

p_ligand_target_network <- make_heatmap_ggplot(vis_ligand_target, "Prioritized CAF-ligands", "p-EMT genes in malignant cells",
                    color = "purple", legend_title = "Regulatory potential") +
  scale_fill_gradient2(low = "whitesmoke",  high = "purple")

p_ligand_target_network

```

We can also look at which receptors of the receiver cell population (malignant cells) can potentially bind to the prioritized ligands from the sender cell population (CAFs).

```{r ligand-receptor-heatmap, fig.width=9, fig.height=6}

ligand_receptor_links_df <- get_weighted_ligand_receptor_links(
  best_upstream_ligands, expressed_receptors,
  lr_network, weighted_networks$lr_sig) 

vis_ligand_receptor_network <- prepare_ligand_receptor_visualization(
  ligand_receptor_links_df,
  best_upstream_ligands,
  order_hclust = "both") 

(make_heatmap_ggplot(t(vis_ligand_receptor_network), 
                     y_name = "Prioritized CAF-ligands", x_name = "Receptors expressed by malignant cells",  
                     color = "mediumvioletred", legend_title = "Prior interaction potential"))

```


## 6. Summary visualizations of the NicheNet analysis

### Load additional packages required for the visualization

```{r}
library(RColorBrewer)
library(cowplot)
library(ggpubr)
```

### Prepare the ligand activity matrix

```{r}
vis_ligand_aupr <- ligand_activities %>% filter(test_ligand %in% best_upstream_ligands) %>%
  column_to_rownames("test_ligand") %>% select(aupr_corrected) %>% arrange(aupr_corrected) %>% as.matrix(ncol = 1)
```

```{r ligand-activity-heatmap, fig.width=5, fig.height=6}
p_ligand_aupr <- make_heatmap_ggplot(vis_ligand_aupr,
                                     "Prioritized CAF-ligands", "Ligand activity",
                                     color = "darkorange", legend_title = "AUPR") + 
    theme(axis.text.x.top = element_blank())
p_ligand_aupr
```

#### Prepare expression of ligands in fibroblast per tumor

Because the single-cell data was collected from multiple tumors, we will show here the average expression of the ligands per tumor.

```{r}
expression_df_CAF <- expression[CAF_ids, best_upstream_ligands] %>% data.frame() %>%
  rownames_to_column("cell") %>% as_tibble() %>%
  inner_join(sample_info %>% select(cell,tumor), by =  "cell")

aggregated_expression_CAF <- expression_df_CAF %>% group_by(tumor) %>%
  select(-cell) %>% summarise_all(mean)

aggregated_expression_df_CAF <- aggregated_expression_CAF %>% select(-tumor) %>% t() %>%
  magrittr::set_colnames(aggregated_expression_CAF$tumor) %>%
  data.frame() %>% rownames_to_column("ligand") %>% as_tibble() 

aggregated_expression_matrix_CAF <- aggregated_expression_df_CAF %>% select(-ligand) %>% as.matrix() %>%
  magrittr::set_rownames(aggregated_expression_df_CAF$ligand)

# This order was determined based on the paper from Puram et al. Tumors are ordered according to p-EMT score.
order_tumors <- c("HN6","HN20","HN26","HN28","HN22","HN25","HN5","HN18","HN17","HN16") 
vis_ligand_tumor_expression <- aggregated_expression_matrix_CAF[rev(best_upstream_ligands), order_tumors]

```

```{r ligand-expression-heatmap, fig.width=9, fig.height=6}
color <- colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100)
p_ligand_tumor_expression <- make_heatmap_ggplot(vis_ligand_tumor_expression,
                                                 "Prioritized CAF-ligands", "Tumor",
                                                 color = color[100],
                                                 legend_title = "Expression\n(averaged over\nsingle cells)")
p_ligand_tumor_expression
```

#### Prepare expression of target genes in malignant cells per tumor

```{r}
expression_df_target <- expression[malignant_ids,geneset_oi] %>% data.frame() %>%
  rownames_to_column("cell") %>% as_tibble() %>%
  inner_join(sample_info %>% select(cell,tumor), by =  "cell") 

aggregated_expression_target <- expression_df_target %>% group_by(tumor) %>%
  select(-cell) %>% summarise_all(mean)

aggregated_expression_df_target <- aggregated_expression_target %>% select(-tumor) %>% t() %>%
  magrittr::set_colnames(aggregated_expression_target$tumor) %>%
  data.frame() %>% rownames_to_column("target") %>% as_tibble() 

aggregated_expression_matrix_target <- aggregated_expression_df_target %>% select(-target) %>%as.matrix() %>%
  magrittr::set_rownames(aggregated_expression_df_target$target)

vis_target_tumor_expression_scaled <- aggregated_expression_matrix_target %>% t() %>% scale_quantile() %>%
  .[order_tumors, order_targets]
```

```{r target-expression-heatmap, fig.width=9, fig.height=6}
p_target_tumor_scaled_expression <- make_threecolor_heatmap_ggplot(vis_target_tumor_expression_scaled,
                                                                   "Tumor", "Target",
                                                                   low_color = color[1], mid_color = color[50], mid = 0.5,
                                                                   high_color = color[100],
                                                                   legend_title = "Scaled expression\n(averaged over\nsingle cells)")
p_target_tumor_scaled_expression

```

#### Combine the different heatmaps in one overview figure

```{r summary-vis, fig.width=15, fig.height=9}

figures_without_legend = plot_grid(
  p_ligand_aupr + theme(legend.position = "none"),
  p_ligand_tumor_expression + theme(legend.position = "none",
                                    axis.title.y = element_blank()),
  p_ligand_target_network + theme(legend.position = "none",
                                  axis.ticks = element_blank(),
                                  axis.title.y = element_blank()), 
  NULL,
  NULL,
  p_target_tumor_scaled_expression + theme(legend.position = "none",
                                           axis.title.x = element_blank()), 
  align = "hv",
  nrow = 2,
  rel_widths = c(ncol(vis_ligand_aupr)+6, ncol(vis_ligand_tumor_expression), ncol(vis_ligand_target))-2,
  rel_heights = c(nrow(vis_ligand_aupr), nrow(vis_target_tumor_expression_scaled)+3)) 

legends = plot_grid(
  as_ggplot(get_legend(p_ligand_aupr)),
  as_ggplot(get_legend(p_ligand_tumor_expression)),
  as_ggplot(get_legend(p_ligand_target_network)),
  as_ggplot(get_legend(p_target_tumor_scaled_expression)),
  nrow = 2,
  align = "h")

plot_grid(figures_without_legend, 
          legends, 
          rel_heights = c(10,2), nrow = 2, align = "hv")


```

```{r}
sessionInfo()
```


## References
