---
title: "NicheNet's ligand activity analysis on a gene set of interest: predict active ligands and their target genes"
author: "Robin Browaeys"
date: "2019-01-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NicheNet's ligand activity analysis on a gene set of interest: predict active ligands and their target genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: library.bib
---

<!-- github markdown built using 
rmarkdown::render("vignettes/ligand_activity_geneset.Rmd", output_format = "github_document")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

In this vignette, you can learn how to perform a basic NicheNet analysis. A NicheNet analysis can help you to generate hypotheses about an intercellular communication process of interest for which you have bulk or single-cell gene expression data. Specifically, NicheNet can predict 1) which ligands from one cell population ("sender/niche") are most likely to affect target gene expression in an interacting cell population ("receiver/target") and 2) which specific target genes are affected by which of these predicted ligands. 

Because NicheNet studies how ligands affect gene expression in neighboring cells, you need to have data about this effect in gene expression you want to study. So, you need to have a clear set of genes that are putatively affected by ligands from one of more interacting cells.

The pipeline of a basic NicheNet analysis consist mainly of the following steps:

* 1. Define a “sender/niche” cell population and a “receiver/target” cell population present in your expression data and determine which genes are expressed in both populations
* 2. Define a gene set of interest: these are the genes in the “receiver/target” cell population that are potentially affected by ligands expressed by interacting cells (e.g. genes differentially expressed upon cell-cell interaction)
* 3. Define a set of potential ligands: these are ligands that are expressed by the “sender/niche” cell population and bind a (putative) receptor expressed by the “receiver/target” population
* 4) Perform NicheNet ligand activity analysis: rank the potential ligands based on the presence of their target genes in the gene set of interest (compared to the background set of genes)
* 5) Infer top-predicted target genes of ligands that are top-ranked in the ligand activity analysis

This vignette guides you in detail through all these steps. As example expression data of interacting cells, we will use data from Puram et al. to explore intercellular communication in the tumor microenvironment in head and neck squamous cell carcinoma (HNSCC) [See @puram_single-cell_2017]. More specifically, we will look at which ligands expressed by cancer-associated fibroblasts (CAFs) can induce a specific gene program in neighboring malignant cells. This program, a partial epithelial-mesenschymal transition (p-EMT) program, could be linked to metastasis by Puram et al. 

The used ligand-target matrix and example expression data of interacting cells can be downloaded from Zenodo. [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3260758.svg)](https://doi.org/10.5281/zenodo.3260758)

## Step 0: Load required packages, NicheNet's ligand-target prior model and processed expression data of interacting cells

Packages:

```{r}
library(nichenetr)
library(tidyverse)
```

Ligand-target model:

This model denotes the prior potential that a particular ligand might regulate the expression  of a specific target gene. 

```{r}
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns
```

Expression data of interacting cells: publicly available single-cell data from CAF and malignant cells from HNSCC tumors:

```{r}
hnscc_expression = readRDS(url("https://zenodo.org/record/3260758/files/hnscc_expression.rds"))
expression = hnscc_expression$expression
sample_info = hnscc_expression$sample_info # contains meta-information about the cells
```

## Step 1: Define expressed genes in sender and receiver cell populations

Our research question is to prioritize which ligands expressed by CAFs can induce p-EMT in neighboring malignant cells. Therefore, CAFs are the sender cells in this example and malignant cells are the receiver cells. This is an example of paracrine signaling. Note that autocrine signaling can be considered if sender and receiver cell type are the same.

Now, we will determine which genes are expressed in the sender cells (CAFs) and receiver cells (malignant cells) from high quality primary tumors. Therefore, we wil not consider cells from tumor samples of less quality or from lymph node metastases. 

To determine expressed genes in this case study, we use the definition used by Puram et al. (the authors of this dataset), which is: Ea, the aggregate expression of each gene i across the k cells, calculated as Ea(i) = log2(average(TPM(i)1…k)+1), should be >= 4. We recommend users to define expressed genes in the way that they consider to be most appropriate for their dataset. For single-cell data generated by the 10x platform in our lab, we don't use the definition used here, but we consider genes to be expressed in a cell type when they have non-zero values in at least 10% of the cells from that cell type. This is described as well in the other vignette [Perform NicheNet analysis starting from a Seurat object: step-by-step analysis](seurat_steps.md):`vignette("seurat_steps", package="nichenetr")`.

```{r}
tumors_remove = c("HN10","HN","HN12", "HN13", "HN24", "HN7", "HN8","HN23")

CAF_ids = sample_info %>% filter(`Lymph node` == 0 & !(tumor %in% tumors_remove) & `non-cancer cell type` == "CAF") %>% pull(cell)
malignant_ids = sample_info %>% filter(`Lymph node` == 0 & !(tumor %in% tumors_remove) & `classified  as cancer cell` == 1) %>% pull(cell)

expressed_genes_sender = expression[CAF_ids,] %>% apply(2,function(x){10*(2**x - 1)}) %>% apply(2,function(x){log2(mean(x) + 1)}) %>% .[. >= 4] %>% names()
expressed_genes_receiver = expression[malignant_ids,] %>% apply(2,function(x){10*(2**x - 1)}) %>% apply(2,function(x){log2(mean(x) + 1)}) %>% .[. >= 4] %>% names()

# Check the number of expressed genes: should be a 'reasonable' number of total expressed genes in a cell type, e.g. between 5000-10000 (and not 500 or 20000)
length(expressed_genes_sender)
length(expressed_genes_receiver)
```

## Step 2: Define the gene set of interest and a background of genes

As gene set of interest, we consider the genes of which the expression is possibly affected due to communication with other cells. The definition of this gene set depends on your research question and is a crucial step in the use of NicheNet.

Because we here want to investigate how CAFs regulate the expression of p-EMT genes in malignant cells, we will use the p-EMT gene set defined by Puram et al. as gene set of interest and use all genes expressed in malignant cells as background of genes. 

```{r}
geneset_oi = readr::read_tsv(url("https://zenodo.org/record/3260758/files/pemt_signature.txt"), col_names = "gene") %>% pull(gene) %>% .[. %in% rownames(ligand_target_matrix)] # only consider genes also present in the NicheNet model - this excludes genes from the gene list for which the official HGNC symbol was not used by Puram et al.
head(geneset_oi)

background_expressed_genes = expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
head(background_expressed_genes)
```

## Step 3: Define a set of potential ligands

As potentially active ligands, we will use ligands that are 1) expressed by CAFs and 2) can bind a (putative) receptor expressed by malignant cells. Putative ligand-receptor links were gathered from NicheNet's ligand-receptor data sources.

```{r}
lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))

# If wanted, users can remove ligand-receptor interactions that were predicted based on protein-protein interactions and only keep ligand-receptor interactions that are described in curated databases. To do this: uncomment following line of code:
# lr_network = lr_network %>% filter(database != "ppi_prediction_go" & database != "ppi_prediction")

ligands = lr_network %>% pull(from) %>% unique()
expressed_ligands = intersect(ligands,expressed_genes_sender)

receptors = lr_network %>% pull(to) %>% unique()
expressed_receptors = intersect(receptors,expressed_genes_receiver)

lr_network_expressed = lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) 
head(lr_network_expressed)
```

This ligand-receptor network contains the expressed ligand-receptor interactions. As potentially active ligands for the NicheNet analysis, we will consider the ligands from this network.

```{r}
potential_ligands = lr_network_expressed %>% pull(from) %>% unique()
head(potential_ligands)
```


## Step 4: Perform NicheNet's ligand activity analysis on the gene set of interest

Now perform the ligand activity analysis: in this analysis, we will calculate the ligand activity of each ligand, or in other words, we will assess how well each CAF-ligand can predict the p-EMT gene set compared to the background of expressed genes (predict whether a gene belongs to the p-EMT program or not).

```{r}
ligand_activities = predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = potential_ligands)
```

Now, we want to rank the ligands based on their ligand activity. In our validation study, we showed that the pearson correlation coefficient (PCC) between a ligand's target predictions and the observed transcriptional response was the most informative measure to define ligand activity. Therefore, we will rank the ligands based on their pearson correlation coefficient. This allows us to prioritize p-EMT-regulating ligands.

```{r}
ligand_activities %>% arrange(-pearson) 
best_upstream_ligands = ligand_activities %>% top_n(20, pearson) %>% arrange(-pearson) %>% pull(test_ligand)
head(best_upstream_ligands)
```

We see here that the performance metrics indicate that the 20 top-ranked ligands can predict the p-EMT genes reasonably, this implies that ranking of the ligands might be accurate as shown in our study. However, it is possible that for some gene sets, the target gene prediction performance of the top-ranked ligands would not be much better than random prediction. In that case, prioritization of ligands will be less trustworthy. 

Additional note: we looked at the top 20 ligands here and will continue the analysis by inferring p-EMT target genes of these 20 ligands. However, the choice of looking only at the 20 top-ranked ligands for further biological interpretation is based on biological intuition and is quite arbitrary. Therefore, users can decide to continue the analysis with a different number of ligands. We recommend to check the selected cutoff by looking at the distribution of the ligand activity values. Here, we show the ligand activity histogram (the score for the 20th ligand is indicated via the dashed line).

```{r}
# show histogram of ligand activity scores
p_hist_lig_activity = ggplot(ligand_activities, aes(x=pearson)) + 
  geom_histogram(color="black", fill="darkorange")  + 
  # geom_density(alpha=.1, fill="orange") +
  geom_vline(aes(xintercept=min(ligand_activities %>% top_n(20, pearson) %>% pull(pearson))), color="red", linetype="dashed", size=1) + 
  labs(x="ligand activity (PCC)", y = "# ligands") +
  theme_classic()
p_hist_lig_activity
```


## Step 5: Infer target genes of top-ranked ligands and visualize in a heatmap

Now we will show how you can look at the regulatory potential scores between ligands and target genes of interest. In this case, we will look at links between top-ranked p-EMT regulating ligands and p-EMT genes. In the ligand-target heatmaps, we show here regulatory potential scores for interactions between the 20 top-ranked ligands and following target genes: genes that belong to the gene set of interest and to the 250 most strongly predicted targets of at least one of the 20 top-ranked ligands (the top 250 targets according to the general prior model, so not the top 250 targets for this dataset). Consequently, genes of your gene set that are not a top target gene of one of the prioritized ligands, will not be shown on the heatmap.

```{r}
active_ligand_target_links_df = best_upstream_ligands %>% lapply(get_weighted_ligand_target_links,geneset = geneset_oi, ligand_target_matrix = ligand_target_matrix, n = 250) %>% bind_rows()

nrow(active_ligand_target_links_df)
head(active_ligand_target_links_df)
```

For visualization purposes, we adapted the ligand-target regulatory potential matrix as follows. Regulatory potential scores were set as 0 if their score was below a predefined threshold, which was here the 0.25 quantile of scores of interactions between the 20 top-ranked ligands and each of their respective top targets (see the ligand-target network defined in the data frame).

```{r}
active_ligand_target_links = prepare_ligand_target_visualization(ligand_target_df = active_ligand_target_links_df, ligand_target_matrix = ligand_target_matrix, cutoff = 0.25)

nrow(active_ligand_target_links_df)
head(active_ligand_target_links_df)
```

The putatively active ligand-target links will now be visualized in a heatmap.
The order of the ligands accord to the ranking according to the ligand activity prediction.

```{r, fig.width=8, fig.height=6}
order_ligands = intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
order_targets = active_ligand_target_links_df$target %>% unique()
vis_ligand_target = active_ligand_target_links[order_targets,order_ligands] %>% t()

p_ligand_target_network = vis_ligand_target %>% make_heatmap_ggplot("Prioritized CAF-ligands","p-EMT genes in malignant cells", color = "purple",legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential") + scale_fill_gradient2(low = "whitesmoke",  high = "purple", breaks = c(0,0.005,0.01)) + theme(axis.text.x = element_text(face = "italic"))

p_ligand_target_network

```

Note that the choice of these cutoffs for visualization is quite arbitrary. We recommend users to test several cutoff values. 

If you would consider more than the top 250 targets based on prior information, you will infer more, but less confident, ligand-target links; by considering less than 250 targets, you will be more stringent. 

If you would change the quantile cutoff that is used to set scores to 0 (for visualization purposes), lowering this cutoff will result in a more dense heatmap, whereas highering this cutoff will result in a more sparse heatmap. 

## Follow-up analysis 1: Ligand-receptor network inference for top-ranked ligands

One type of follow-up analysis is looking at which receptors of the receiver cell population (here: malignant cells) can potentially bind to the prioritized ligands from the sender cell population (here: CAFs).

So, we will now infer the predicted ligand-receptor interactions of the top-ranked ligands and visualize these in a heatmap.

```{r}
# get the ligand-receptor network of the top-ranked ligands
lr_network_top = lr_network %>% filter(from %in% best_upstream_ligands & to %in% expressed_receptors) %>% distinct(from,to)
best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()

# get the weights of the ligand-receptor interactions as used in the NicheNet model
weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
lr_network_top_df = weighted_networks$lr_sig %>% filter(from %in% best_upstream_ligands & to %in% best_upstream_receptors)

# convert to a matrix
lr_network_top_df = lr_network_top_df %>% spread("from","weight",fill = 0)
lr_network_top_matrix = lr_network_top_df %>% select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df$to)

# perform hierarchical clustering to order the ligands and receptors
dist_receptors = dist(lr_network_top_matrix, method = "binary")
hclust_receptors = hclust(dist_receptors, method = "ward.D2")
order_receptors = hclust_receptors$labels[hclust_receptors$order]

dist_ligands = dist(lr_network_top_matrix %>% t(), method = "binary")
hclust_ligands = hclust(dist_ligands, method = "ward.D2")
order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]
```

Show a heatmap of the ligand-receptor interactions

```{r, fig.width=9, fig.height=6}
vis_ligand_receptor_network = lr_network_top_matrix[order_receptors, order_ligands_receptor]
p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Prioritized CAF-ligands","Receptors expressed by malignant cells", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential")
p_ligand_receptor_network

```

## Follow-up analysis 2: Visualize expression of top-predicted ligands and their target genes in a combined heatmap

NicheNet only considers expressed ligands of sender cells, but does not take into account their expression for ranking the ligands. The ranking is purely based on the potential that a ligand might regulate the gene set of interest, given prior knowledge. Because it is also useful to further look into expression of ligands and their target genes, we demonstrate here how you could make a combined figure showing ligand activity, ligand expression, target gene expression and ligand-target regulatory potential. 

#### Load additional packages required for the visualization:

```{r}
library(RColorBrewer)
library(cowplot)
library(ggpubr)
```

#### Prepare the ligand activity matrix

```{r}
ligand_pearson_matrix = ligand_activities %>% select(pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities$test_ligand)

vis_ligand_pearson = ligand_pearson_matrix[order_ligands, ] %>% as.matrix(ncol = 1) %>% magrittr::set_colnames("Pearson")
```

```{r, fig.width=5, fig.height=6}
p_ligand_pearson = vis_ligand_pearson %>% make_heatmap_ggplot("Prioritized CAF-ligands","Ligand activity", color = "darkorange",legend_position = "top", x_axis_position = "top", legend_title = "Pearson correlation coefficient\ntarget gene prediction ability)")
p_ligand_pearson
```

#### Prepare expression of ligands in fibroblast per tumor

Because the single-cell data was collected from multiple tumors, we will show here the average expression of the ligands per tumor.

```{r}
expression_df_CAF = expression[CAF_ids,order_ligands] %>% data.frame() %>% rownames_to_column("cell") %>% as_tibble() %>% inner_join(sample_info %>% select(cell,tumor), by =  "cell")

aggregated_expression_CAF = expression_df_CAF %>% group_by(tumor) %>% select(-cell) %>% summarise_all(mean)

aggregated_expression_df_CAF = aggregated_expression_CAF %>% select(-tumor) %>% t() %>% magrittr::set_colnames(aggregated_expression_CAF$tumor) %>% data.frame() %>% rownames_to_column("ligand") %>% as_tibble() 

aggregated_expression_matrix_CAF = aggregated_expression_df_CAF %>% select(-ligand) %>% as.matrix() %>% magrittr::set_rownames(aggregated_expression_df_CAF$ligand)

order_tumors = c("HN6","HN20","HN26","HN28","HN22","HN25","HN5","HN18","HN17","HN16") # this order was determined based on the paper from Puram et al. Tumors are ordered according to p-EMT score.
vis_ligand_tumor_expression = aggregated_expression_matrix_CAF[order_ligands,order_tumors]
```

```{r, fig.width=9, fig.height=6}
library(RColorBrewer)
color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100)
p_ligand_tumor_expression = vis_ligand_tumor_expression %>% make_heatmap_ggplot("Prioritized CAF-ligands","Tumor", color = color[100],legend_position = "top", x_axis_position = "top", legend_title = "Expression\n(averaged over\nsingle cells)") + theme(axis.text.y = element_text(face = "italic"))
p_ligand_tumor_expression
```

#### Prepare expression of target genes in malignant cells per tumor

```{r}
expression_df_target = expression[malignant_ids,geneset_oi] %>% data.frame() %>% rownames_to_column("cell") %>% as_tibble() %>% inner_join(sample_info %>% select(cell,tumor), by =  "cell") 

aggregated_expression_target = expression_df_target %>% group_by(tumor) %>% select(-cell) %>% summarise_all(mean)

aggregated_expression_df_target = aggregated_expression_target %>% select(-tumor) %>% t() %>% magrittr::set_colnames(aggregated_expression_target$tumor) %>% data.frame() %>% rownames_to_column("target") %>% as_tibble() 

aggregated_expression_matrix_target = aggregated_expression_df_target %>% select(-target) %>% as.matrix() %>% magrittr::set_rownames(aggregated_expression_df_target$target)

vis_target_tumor_expression_scaled = aggregated_expression_matrix_target %>% t() %>% scale_quantile() %>% .[order_tumors,order_targets]
```

```{r, fig.width=9, fig.height=6}
p_target_tumor_scaled_expression = vis_target_tumor_expression_scaled  %>% make_threecolor_heatmap_ggplot("Tumor","Target", low_color = color[1],mid_color = color[50], mid = 0.5, high_color = color[100], legend_position = "top", x_axis_position = "top" , legend_title = "Scaled expression\n(averaged over\nsingle cells)") + theme(axis.text.x = element_text(face = "italic"))
p_target_tumor_scaled_expression

```

#### Combine the different heatmaps in one overview figure

```{r, fig.width=13, fig.height=7}
figures_without_legend = plot_grid(
  p_ligand_pearson + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()),
  p_ligand_tumor_expression + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()) + ylab(""),
  p_ligand_target_network + theme(legend.position = "none", axis.ticks = element_blank()) + ylab(""), 
  NULL,
  NULL,
  p_target_tumor_scaled_expression + theme(legend.position = "none", axis.ticks = element_blank()) + xlab(""), 
  align = "hv",
  nrow = 2,
  rel_widths = c(ncol(vis_ligand_pearson)+ 4.5, ncol(vis_ligand_tumor_expression), ncol(vis_ligand_target)) -2,
  rel_heights = c(nrow(vis_ligand_pearson), nrow(vis_target_tumor_expression_scaled) + 3)) 

legends = plot_grid(
  as_ggplot(get_legend(p_ligand_pearson)),
  as_ggplot(get_legend(p_ligand_tumor_expression)),
  as_ggplot(get_legend(p_ligand_target_network)),
  as_ggplot(get_legend(p_target_tumor_scaled_expression)),
  nrow = 2,
  align = "h")

plot_grid(figures_without_legend, 
          legends, 
          rel_heights = c(10,2), nrow = 2, align = "hv")


```

## Other follow-up analyses:

As another follow-up analysis, you can infer possible signaling paths between ligands and targets of interest. You can read how to do this in the following vignette [Inferring ligand-to-target signaling paths](ligand_target_signaling_path.md):`vignette("ligand_target_signaling_path", package="nichenetr")`. 

Another follow-up analysis is getting a "tangible" measure of how well top-ranked ligands predict the gene set of interest and assess which genes of the gene set can be predicted well. You can read how to do this in the following vignette [Assess how well top-ranked ligands can predict a gene set of interest](target_prediction_evaluation_geneset.md):`vignette("target_prediction_evaluation_geneset", package="nichenetr")`. 

In case you want to visualize ligand-target links between multiple interacting cells, you can make an appealing circos plot as shown in vignette [Circos plot visualization to show active ligand-target links between interacting cells](circos.md):`vignette("circos", package="nichenetr")`.

## References
