---
title: "Evaluation of NicheNet's ligand-target predictions"
author: "Robin Browaeys"
date: "2018-11-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This vignette shows how the ligand-target predictions of NicheNet were evaluated. For validation, we collected transcriptome data of cells before and after they were treated by one or two ligands in culture. Using these ligand treatment datasets for validation has the advantage that observed gene expression changes can be directly attributed to the addition of the ligand(s). Hence, differentially expressed genes can be considered as a gold standard of target genes of a particular ligand. 

You can use the procedure shown here to evaluate your own model and compare its performance to NicheNet. Ligand treatment validation datasets and NicheNet's ligand-target model can be downloaded from Zenodo [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1484138.svg)](https://doi.org/10.5281/zenodo.1484138). 


### Load nichenetr, the model we want to evaluate, and the datasets on which we want to evaluate it.
```{r}
library(nichenetr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(tibble)

# Load in all expression datasets
expression_settings_validation = readRDS(url("https://zenodo.org/record/1484138/files/expression_settings.rds"))

# Load in the ligand-target model
ligand_target_matrix = readRDS(url("https://zenodo.org/record/1484138/files/ligand_target_matrix.rds"))

```

### Example: transcriptional response prediction evaluation

First, we will demonstrate how to evaluate the transcriptional response (i.e. target gene prediction) performance for all ligand treatment expression datasets. For this, we determine how well the model predicts which genes are differentially expressed after treatment with a ligand. Ideally, target genes with high regulatory potential scores for a ligand, should be differentially expressed in response to that ligand.
```{r, fig.width=8, fig.height=8}
# Convert expression datasets to the required format to perform target gene prediction
settings = expression_settings_validation %>% lapply(convert_expression_settings_evaluation)

# For the sake of simplicity, we exclude in this vignette the ligand-treatment datasets profiling the response to multiple ligands. To see how to build a ligand-target model with target predictions for multiple ligands at once: see vignette [Construction of NicheNet's ligand-target model](vignettes/model_construction.md): `vignette("model_construction", package="nichenetr")`.
settings = settings %>% discard(~length(.$from) > 1)

# Evaluate transcriptional response prediction on every dataset
performances = settings %>% lapply(evaluate_target_prediction, ligand_target_matrix) %>% bind_rows()

# Visualize some classification evaluation metrics showing the target gene prediction performance
performances = performances %>% select(-aupr, -auc_iregulon, -sensitivity_roc, -specificity_roc) %>% gather(key = scorename, value = scorevalue, auroc:spearman)
scorelabels = c(auroc="AUROC", aupr_corrected="AUPR (corrected)", auc_iregulon_corrected = "AUC-iRegulon (corrected)",pearson = "Pearson correlation", spearman = "Spearman's rank correlation",mean_rank_GST_log_pval = "Mean-rank gene-set enrichment")
scorerandom = c(auroc=0.5, aupr_corrected=0, auc_iregulon_corrected = 0, pearson = 0, spearman = 0,mean_rank_GST_log_pval = 0) %>% data.frame(scorevalue=.) %>% rownames_to_column("scorename")

performances %>% 
  mutate(model = "NicheNet") %>% 
  ggplot() + 
  geom_violin(aes(model, scorevalue, group=model, fill = model)) + 
  geom_boxplot(aes(model, scorevalue, group = model),width = 0.05) + 
  scale_y_continuous("Score target prediction") + 
  facet_wrap(~scorename, scales = "free", labeller=as_labeller(scorelabels)) +
  geom_hline(aes(yintercept=scorevalue), data=scorerandom, linetype = 2, color = "red") +
  theme_bw()

```

### Example: ligand activity prediction evaluation

Now we will show how to assess the accuracy of the model in predicting whether cells were treated by a particular ligand or not. In other words, we will evaluate how well NicheNet prioritizes active ligand(s), given a set of differentially expressed genes. For this procedure, we assume the following: the better a ligand predicts the transcriptional response compared to other ligands, the more likely it is that this ligand is active. Therefore, we first get ligand activity/importance scores for all ligands on all ligand-treatment expression datasets of which the true acive ligand is known. Then we assess whether the truly active ligands get indeed higher ligand activity scores as should be for a good ligand-target model.

```{r,fig.width=8, fig.height=8}
# convert expression datasets to correct format for ligand activity prediction
all_ligands = settings %>% extract_ligands_from_settings(combination = FALSE) %>% unlist()
settings_ligand_prediction = settings %>% convert_settings_ligand_prediction(all_ligands = all_ligands, validation = TRUE)

# infer ligand importances: for all ligands of interest, we assess how well a ligand explains the differential expression in a specific datasets (and we do this for all datasets).
ligand_importances = settings_ligand_prediction %>% lapply(get_single_ligand_importances,ligand_target_matrix) %>% bind_rows()

# Look at predictive performance of single/individual importance measures to predict ligand activity: of all ligands tested, the ligand that is truly active in a dataset should get the highest activity score (i.e. best target gene prediction performance)
evaluation_ligand_prediction = ligand_importances %>% evaluate_single_importances_ligand_prediction(normalization = "median")

# Visualize some classification evaluation metrics showing the ligand activity prediction performance
evaluation_ligand_prediction = evaluation_ligand_prediction %>% select(-aupr, -sensitivity_roc, -specificity_roc, -pearson, -spearman, -mean_rank_GST_log_pval) %>% gather(key = scorename, value = scorevalue, auroc:aupr_corrected)
scorelabels = c(auroc="AUROC", aupr_corrected="AUPR (corrected)")
scorerandom = c(auroc=0.5, aupr_corrected=0) %>% data.frame(scorevalue=.) %>% rownames_to_column("scorename")

evaluation_ligand_prediction %>% 
 filter(importance_measure %in% c("auroc", "aupr_corrected", "mean_rank_GST_log_pval", "auc_iregulon_corrected", "pearson", "spearman")) %>%
  ggplot() + 
  geom_point(aes(importance_measure, scorevalue, group=importance_measure, color = importance_measure), size = 3) + 
  scale_y_continuous("Evaluation ligand activity prediction") + 
  scale_x_discrete("Ligand activity measure") + 
  facet_wrap(~scorename, scales = "free", labeller=as_labeller(scorelabels)) +
  geom_hline(aes(yintercept=scorevalue), data=scorerandom, linetype = 2, color = "red") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```
